---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# And finally some code for the re-tracing experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
# And finally, code to compare this data with Sirui's PSEs
source('R/time.R')
```

_1: Centre for Vision Research, York University, Toronto, ON M3J1P3, Canada_\
_2: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
_3: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\

*: Corresponding author: \
Marius ’t Hart\
Centre for Vision Research\
York University\
Toronto, ON M3J1P3\
Canada\
thartbm@gmail.com\


# Abstract

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006]. The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously track the position of the perceived position of the stimulus, using a stylus on a drawing tablet. While the orientation of the recorded trajectories reflect stimulus properties, there were no spontaneous resets. In a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

# Introduction

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence its perceived location, making it appear to be slightly ahead of its true location, e.g. the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013]. These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. And although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018]. These findings suggest that both the true and illusory positions are available in the brain.

```{r fig1, echo=F, fig.height=8, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Top left:_ a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path. Eye-movements immediately following disappearance of the gabor land at the true position, but pointing movements and delayed eye-movements go to the illusory position. _Bottom left:_ at some point, the perceived position will 'reset' towards the true position. For some people this takes the form of a 'jump' back to the true position, for others it means the illusory position doesn't deviate more from the true position, and there are some in between version. _Top right:_ if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate. _Middle right:_ if they occur after some period of time, they should have a constant Y coordinate. _Bottom right:_ we have people track or re-trace their percept and try to isolate reset points from the drawn trajectories to figure out where or when resets occur."}
library(magick)
Fig1 <- image_read_svg('doc/stimulus_reset.svg')
par(mai=c(0,0,0,0))
plot(Fig1)
```


The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left) or it saturates. This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory. Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets. Alternatively, there is a simple periodicity or stochastics process underlies the changes in perception.

Here, we examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing. However, recent work has suggested that movement-related attention may lead to resets [@Nakayama2020], so that continuous real-time tracking of the gabor's position, may prevent the illusory shift in position. The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. Hence, we first test if the illusion occurs when people continunously track the position of the double drift stimulus. Then we try to capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus while seeing it (a tracking procedure), or after having seen it (a retrace procedure). Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion or a temporal limit.

# Methods

## Participants

For the first task in experiment 1, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75). For the second task in experiment 1, 6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6). For experiment 2, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported to be right handed and had either normal or corrected to normal vision. Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.

## Setup

Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses. Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet. The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus.

For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assumed the participants to be close to the setup, with their eyes at the height of the monitor. Since there was no requirement to use the forehead rest, the latter is probably only true for the tallest participants, so that reported _dva's_ are slightly underestimating the size of stimuli.

## Stimuli

The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ ( or ~.6 _dva_ when in the centre of the screen and viewed from a typical distance ). The whole stimulus extended 3 times the sd of the Guassian in all four cardinal directions. The stimulus used 1.5 phases of a sine wave, so that each sine phase equals ~0.58 _cm_ or ~1.5 _dva_. By applying both real motion and phases changes of the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path. The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_ ). The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display. This real motion was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that stimulus would move from either extreme to the other ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ). For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).

## Tasks

In the two tasks in experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus. In experiment 2, participants first watched a double-drift stimulus and then either reproduce the perceived path of the stimulus by retracing it on the tablet, or indicate the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus. More details are given below.

## Analyses

For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second ( leaving 1 _s_ trajectories for the first experiment, and 3 _s_ trajectories for the second experiment). We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples). This heading measure should depend on the illusion strength. Moreoever, when spontaneous resets are either time-dependent or space-dependent, resets should occur at roughly the same points and in the same direction in a given condition for every participant. This means that spontaneous resets should result in systematic deviations of every participant's average heading over time.

For comparison, we use the same heading measure in experiment 2. However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect direction changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 _cm_ away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.

Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].

# Experiment 1: Continuous tracking

In this experiment, two tasks were performed by two separate groups of participants. Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion. The idea was to test both if tracking the illusion continuously would preserve the illusion and if the tracked paths reveal resets.

## Tasks

In the first task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so. Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted. Internal motion could be 3, 1, 0, -1 or -3 _cps_. Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path.

In the second task, only the highest internal motion was used (±3 _cps_), trials include only 3 complete passes, but these took 4 _s_ each. As before, the fixation point (left or right) was random for each trial, while initial external motion (forward or backward) as well as internal motion were counterbalanced across trials.

## Results

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking preserves the illusion. We will use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.

```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2. Heading and Trajectories in Experiment 1a.** **Top:** Average heading across the middle 1 _s_ of each pass. Distance from origin is time, angle is angle. In blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading. **Bottom:** All individual trajectores. _Red_: Highlighted trajectories from conditions with non-zero internal speed, where at least some heading samples fell in the 95% confidence interval for that participants' heading distribution in zero internal speed condition."}
plotBoundedTracking()
```

The average headings are close to straight lines for all conditions. Since average straight trajectories could either result from uniform tracking paths, each with no signs of resets or from paths with many differently shaped, directed and timed resets, they do not prove or disprove neither the absence nor the presence of resets. Hence, we also show full trajectories of all participants in each of the 5 internal motion conditions.

In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus. Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition. We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset. These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.

In a second task, we only used the higher internal speed (±3 _cps_) and increased the duration of a single pass of the stimulus from 2 _s_ to 4 _s_, both the increase the occurence of spontaneous resets. Average headings and heading distributions for 6 participants can be seen in Figure 3.

```{r fig3, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig3}**Figure 3. Heading in Experiment 1b.** Average heading and heading distribution across the middle 3 _s_ of each pass. Distance from origin is time, angle is angle. In dark blue lines the average heading, in light blue a polar heatmap giving an impression of the distribution of heading. **Left:** Data for all participants. **Right:** Group average."}
plotBoundedTracking_V2()
```

Except for the first part of participant 12's trajectories, all heading data matched that of the 3 _cps_ condition in experiment 1a. Those first few samples in participant 12 could reflect a participant specific delay in tracking the stimulus position. In other words, no unambiguous resets were captured in either experiment. It is possible that resets were perceived but were too quick to respond to in the continuous tracking task. Alternatively, tracking may have stabilized the illusion, preventing resets from occurring. Hence, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.

# Experiment 2: Re-tracing

In a second experiment, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus. This allows participants more time to carefully reproduce the path. We use the recorded trajectories to assess points where resets occured. The reset points can then be used to determine if resets are time-limited or space-limited.

## Tasks

Here we used a re-tracing procedure where participants drew the perceived path of a double-drift stimulus making a single pass across the workspace. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. Nakayama & Holcombe [-@Nakayama2020] have also used a re-tracing paradigm, in their case to study triggered resets. The re-tracing task was done in half of 8 blocks. In the other half, participants oriented a line to indicate the initial perceived direction of the stimulus motion. Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times. Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.

In both kinds of trials, participants first had to move the stylus to the start position of the gabor, and then fixate a point to the left or right of the centre of the workspace. The double-drift gabor started at the near end of the display, and moved along the horizontal display away from the participant, always making a single, forward pass. After, the gabor had disappeared, participants could respond in one of two ways. When re-tracing the perceived path with the stylus on the tablet, the drawn path would show up as a red line. Participants could "key in" their response by pressing ENTER, or start over by pressing ESCAPE. When indicating only the initial direction of movement, participants would orient a line ( 75 _px_ or ~3 _cm_ ) originating from the starting point of the gabor’s path. They could change the orientation by pressing either the LEFT ARROW key to orient the line more clockwise or the RIGHT ARROW key to orient the line more counter clockwise. They could "key in" their response by pressing ENTER. In both tasks, there was no time-limit for the response, and careful responding was encouraged.

Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, and the depence on the strength of the illusion could be checked with the data from the line-orienting task. 

## Results

In Figure 4, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants. While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4. Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
plotExampleData()
```

On average, the algorithm detects resets in 75% of trials. For three participants, only about half the trials show resets (42%, 44% and 58%). For the remaining six participants the algorithm detects resets in 72%-99% of trials, with an average of 88%. The minimum number of resets in one of the six conditions for any participant is 7 (out of 24), with the average at 17. Since participants experiencing a "hit-the-wall reset" could display a direction change in their paths only by chance, 50% of trials with resets seems reasonable. Neither percept is incorrect, and the average reset positions still indicates some limit to the illusion. Therefore, we keep all participants and resets.

First, we consider the line angle responses as a measure of illusion strength. The strength of the illusion should depend on both the internal and external speed of the double drift stimulus. As can be seen in Figure 5 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles. To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors. We find main effects of both internal speed ( _F_(2,16)=12.7; _p_<.001, $\eta^2$=0.049) and external speed ( _F_(1,8)=6.89; _p_=.03; $\eta^2$=0.031), and no interaction. That is: we verify that illusion strength varied systematically with our manipulations of the stimulus.

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5. Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength. **Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions. Shaded areas denote 95% confidence intervals. **Right:** Angles of samples halfway to a reset over line angle response. Each dot corresponds to one participants average in one of the 6 conditions. Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
plotValidity()
```

We also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength. If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related. As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity. We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$). Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength. From here on, we will only use the angle of samples half-way to detected reset points as measure of illusion strength.

## Single Limits on the Double-Drift Illusion

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 6 (left), we can see that the reset points vary more along the Y-axis than the X-axis. That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff). In fact, the standard deviation of reset point X-coordinates is about 21.9% of teh standard deviation of reset point Y-coordinates. This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.

```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6. Reset Points and Single Limits** **Left:** The average reset points for each participant and stimulus condition displayed spatially. Y-coordinates have been converted to time, in order to normalize for the two external speeds. Transparent purple lines indicate the standard deviation of coordinates. Dashed vertical red line indicates the median X-coordinate (1.6 _cm_). Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_). **Top right:** X-coordinates predicted from a single temporal limit. **Bottom right:** Y-coordinates predicted from a single spatial limit."}
plotResetPoints()
```

Nevertheless, we test two simple models, based on straighforward trigonometry. First, we calculate the slope of a line through the origin and a sample halfway to the reset point. Then, if the perceived path of the gabor falls on a straight line, and reset points occur at a set limit, then a temporal limit would predict X coordinates:

$x = L_y / s$

And a spatial limit would predict Y coordinates:

$y = L_x \cdot s$

Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the slope. If we fit these simple models by minimizing the mean square error (as the distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=258.4) and a spatial limit of ~1.7 _cm_ (AIC=217.7). This is close to the median values of the respective coordinates. The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs). This again indicates that a spatial limit explains more of the data.

However, it may be that after crossing a spatial limit, it takes some given amount of time before a reset of the illusion is perceived. This would mean that both limits explain separate parts of the data. We model this by finding optimal parameter values for both a spatial limit and a temporal limit simultaneously, when minimizing the distance between predicted and atcual reset points (in centimeters). The equations in the model are:

$x = L_x + L_y / s$

$y = L_x \cdot s + L_y$

Figure 7 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).

```{r fig7, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Combined Limits**"}
plotSequentialModel()
```

The limits this model generates are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit. While the MSE and AIC of the model are lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53). This means that a simple spatial limit accounts for most of the data. It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may follow a poisson distribution, which we do not account for here. However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

```{r comment, echo=FALSE, message=FALSE}
#It may be that during online tracking, proprioceptive information on the location of the hand is included in the estimate of the location of the stimulus. This may have decreased the likelihood of a reset, and may also changed the strength of the illusion. Since some participants did both experiments (X/9), we can at least test if illusion strength is different for online tracking and offline re-tracing.
```

## Comparison with PSE data

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. In an as of yet unpublished study [Sirui] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared. The double-drift stimulus was on a 2 or 4 DVA length trajectory, traversed in 1, 2 or 3 seconds. The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7%/s), that is: it explains the data using only a temporal bound on the illusion. Interestingly, this fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. We will compare the two data sets here.

```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
plotAverageTrajectories()
```

If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2. In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials should also be included in the PSE data. Further, we scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus. Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace. For the 4-second stimuli, we removed the last second. Figure 8 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.

For all internal motion speeds, we see a very similar average trajectory, that does not get closer to veridical over time. We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 cycles/s) and time (1, 2, or 3 seconds after stimulus onset) as factors. If the trajectory data from experiment 2 here follows the same pattern as the PSE data, there should be an effect of time, and perhaps an interaction with internal motion speed. We find no effect of time ( _F_(2,16)=0.011, _p_=.957, $\eta^2$<.001), no effect of internal motion speed ( _F_(2,16)=0.127, _p_=.814, $\eta^2$=.0015) and no interaction ( _F_(4,32)=0.361, _p_=.635, $\eta^2$=.002). This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. First, we notice that the trajectory data is highly dependent on participants. It could be that the re-tracing method provides very noisy estimates, hiding any effect. It could also be that the effects truly vary a lot between participants. Since we can see that some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position, it may that the differences between the data sets can simply be explained by random sampling of relatively small numbers of participants.

# Discussion

We set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli. To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus online. While the heading of the tracking movements is sensitive to the strength of the illusion, there are no clear reset points in the tracking data. In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass. In this re-tracing experiment the drawn traces reflect the strength of the illusion just as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset. The positions of these reset points do no vary much with elapsed time, but are better explained by the distance from the true position. This has implications for the way position perception relies on motion signals.

There is a generative model, based on Bayesian cue integration, that can explain many phenomena with motion induced position shifts. This model, formulated by Kwon, Tadin & Knill [-@Kwon2015], is compatbile with a good portion of our data. The model posits there are three signals contributing to an object's perceived position: it's global motion, it's local motion (or texture motion) and it's position. Especially when the directly perceived position is a noisy or weak signal, the two motion signals contribute more, as in the double-drift or curveball illusion. In this stimulus, the model predicts that the sensory motion signals bias perceived position away from the true location, but at some point this is balanced by the weak, noisy sensory position signal that bias perceived position back towards the true location [@Kwon2015, Fig 3]. When this balance is reached the perceived path of the object would be parallel to but somewhat displaced from the true object path. This pattern of perceived paths is exhibited by some of our participants, and we have called these "hit-the-wall resets" here.

Some of our participants seem to perceive "jump resets" though, and the model as published, does not predict any such resets. This could be added by including a causal inference step [e.g. @Wei2011]. For the subset of participants that experience jump resets, there may be a process that compares the two position signals (perceived and retinal) and when these exceed some threshold, decides that the two position signals do not share a common source. At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again. For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or there may be another reason they do not employ any causal inference process.

Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2020], particularly "jump" resets. In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea. Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass. However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead stimulus triggered. This should however be tested with eye-tracking data, so that this will remain a speculation here.

## Conclusion

We find that online, manual tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets. We do find spontaneous resets in delayed, single-pass re-tracing. The illusion appears to reset, or max out, predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true (retinal) and illusory (perceived) position of the stimulus. This confirms the notion that we gauge the position of objects in the world by combining both retinal position signals as well as accumulated position shifts derived from motion signals.

# Acknowledgements

The research was support by funding from NSERC Canada (PC and DYPH), as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).

# References {#references .unnumbered}