---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
bibliography: bibliography.bib
csl: journal-of-vision.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# And finally some code for the re-tracing experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
```

_1: Centre for Vision Research, York University, Toronto, ON M3J1P3, Canada_\
_2: Department of Psychology, Glendon College, Toronto, ON M4N3M6, Canada_\
_3: Department of Psychological and Brain Sciences, Dartmouth College, Hanover, NJ 03755, USA_\

*: Corresponding author: \
Marius ’t Hart\
Centre for Vision Research\
York University\
Toronto, ON M3J1P3\
Canada\
thartbm@gmail.com\


# Abstract

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006]. The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously (for XX min) track the position of the perceived position of the stimulus. While the orientation of the trajectories reflect stimulus properties, there were no spontaneous nor any evoked resets.  That’s why, in a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

# Introduction

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence it perceived location, making it appear to be slightly ahead of its true location (e.g. the flash lag, @Nijhawan1994; or the flash grab effect, @Cavanagh2013). These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. This result indicates that both true and illusory locations are available in the brain. Indeed, although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018].


```{r echo=F, fig.height=8, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Top left:_ a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path. Eye-movements immediately following disappearance of the gabor land at the true position, but pointing movements and delayed eye-movements go to the illusory position. _Bottom left:_ at some point, the perceived position will 'reset' towards the true position. For some people this takes the form of a 'jump' back to the true position, for others it means the illusory position doesn't deviate more from the true position, and there are some in between version. _Top right:_ if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate. _Middle right:_ if they occur after some period of time, they should have a constant Y coordinate. _Bottom right:_ we have people track or re-trace their percept and try to isolate reset points from the drawn trajectories to figure out where or when resets occur."}
library(magick)
Fig1 <- image_read_svg('doc/stimulus_reset.svg')
par(mai=c(0,0,0,0))
plot(Fig1)
```


The strength of the double-drift probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location to reach deviations of several times the size of the gabor. At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left). This long duration before resetting is well beyond the integration period of other visual effects and into the range of short-term memory. Here, we first examine the double-drift illusion by asking participants to trace out its path with a stylus on a tablet that is optically aligned with the monitor surface but with the hand not visible. Earlier results for pointing to the double-drift target [@Lisi2017] show that the illusion persists for manual pointing. However, with online tracking, the presence of the hand near the location of the perceived gabor may interact with the illusion.

The ability to manually track the perceived position will also allow us to study the resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. It could be that the imprecise, but true, retinal position signal limits the illusion, or it could be that – like other illusions, such as bi-stable stimuli – there is a simple periodicity underlying changes in perception. Here we capture where and when observers experience resets by having them draw the perceived trajectory of the stimulus after having seen it (a retrace procedure). This will allow us to distinguish between a spatial limit to the illusion or a temporal limit.

# Methods

## Participants

For the first task in experiment 1, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75). For the second task in experiment 1, 6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6). For experiment 2, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported to be right handed and have either normal or corrected to normal vision.  Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent. 

## Setup

Participants used a stylus on a horizontal drawing tablet (Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1e), as well as a keypad for additional responses. Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), level with the drawing tablet. These were observed through an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand. The stimuli were phase-changing gabor patches 
- size of stimuli in cm, speed in cm/s (with guesstimates of dva?)

## Analyses

Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].


# Experiment 1: Continuous tracking
In this experiment, two tasks were performed by two separate groups of participants. Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion. The idea was that whenever spontaneous resets occur, this would be recorded in the tracking data.

## Tasks

Two similar tasks were used in this experiment. In the first task, 5 different amounts of internal motion were used, including no internal motion as a baseline (-3, -1, 0 , 1 and 3 cps). The stimulus started in the middle, and moved forward and backward several times including 5 complete passes of 13.5 cm (~XXX dva) and 2 seconds each. Each stimulus was repeated 8 times, with randomized position of the fixation point (left or right) of the stimulus and counterbalanced initial external motion direction (forward or backward). This task allowed testing how sensitive online tracking is to the illusion, and served to gauge what resets would look like in tracking data.

In the second task, only the highest internal motion was used (-3 or 3 cps), trials include only 3 complete passes, but these took 4 seconds each. As before, the fixation point (left or right) was random for each trial and initial external motion (forward or backward) was counterbalanced across trials.

## Results

In a first experiment, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of X seconds, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen (where the gabor would appear), and was then asked to fixate a point on the left or right of the screen, and given X seconds to do so. Since the screen was seen via a mirror, the stimulus appeared to move on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far ends of the external motion, and could be X, Y, 0, -Y or -X. Since each moving gabor was shown for Z seconds, it completed A full and B half sweeps of the path (or C cycles), which should allow for spontaneous resets to occur frequently.

Furthermore, since a short disappearance of the stimulus (~250 ms or more?) has been shown to evoke a reset (XXXX; XXXX), these were included in some of the stimuli (X/Y). Each stimulus had either 0 disappearances of the stimulus (X%), or one in the second (X%), third(X%) or fourth (X%) full sweep, and it would either occur at 25% or 75% of the sweep, but never close to the middle where the perceived and illusory position should be the same. This allows characterizing what a forced reset looks like in the trajectories, and then detect spontaneous resets, that still have a lot of opportunity to occur.

In Figure X, we plot the full trajectories of all participants in each of the 5 conditions (internal motion), as well as the distribution and average instantaneous heading direction during the middle second of each 2-second pass of the stimulus. In the 0 internal motion condition the physical and perceived path coincide and there is no illusion. Trajectories from other conditions with heading in the 0 internal motion confidence interval may be evidence of a reset. Trajectories where this occurs are drawn on top of the heading distribution for the group. The average angle of the tracked path is superimposed as both a line, and a numeral (in degrees deviation from straight forward). Both from the raw trajectories as well as from the averaged angles it becomes clear that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a factor (F(df)=XXX, p=XXX, eta=XXX). This means we can use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion for a subset of participants that also took part in further tasks.

However, only a few potential resets / no clear resets were captured by the continuous tracking data. It is possible that resets were perceived but were too quick to respond to in the continuous drawing. Alternatively, since the hand was at the positioned at the perceived location of the gabor this may have provided an additional position signal that stabilized the illusion, preventing resets from occurring. Hence, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.

# Experiment 2: Re-tracing

In a second experiment, we used a retracing where the participants drew what they saw of a single pass trajectory after it was over. This provided more time for the participants to reproduce the resets if there were any – without having to trace them in real time. The double-drift gabor started at the bottom of the display, close to the participant, and moved forward along the horizontal display away from the participant. There were two response conditions. In one, the participant re-traced the path they saw the stimulus take, while in the other, they indicated the angle of the first part of the stimulus motion by orienting a line originating from the starting point of the gabor’s path. The latter has already been done [XXXX] so this allows comparison of the re-tracing responses with earlier work. In the meantime, Nakayama & Holcombe [XXXX] have also used a re-tracing paradigm to study triggered resets.

## Tasks

Participants alternated between blocks of two kinds of trials (in counterbalanced order). In both types of trial, the gabor was shown starting close to the participant and moving away in a forward direction, for a single pass along the workspace, taking 4 seconds. During presentation of the gabor, the participant was instructed to maintain fixation to the left or right of the stimulus. After the gabor disappeared, the participant could break fixation, and was to report the percept in one of two ways. In one kind of block, the participant re-traced the perceived path of the gabor using the stylus. The drawn path was shown as a line on the screen. Participants could “key in” their re-traced percept by pressing ENTER or start over by pressing the ESCAPE key. In the other kind of block, participants reported the initial direction of the gabor (before any resets occurred), by using the LEFT and RIGHT arrow keys to change the orientation of a line of XX cm, with it’s origin in the same spot as where the gabor started moving. Participants could change the line’s orientation it until satisfied and then “key in” their response by pressing ENTER. The arrow report task has been used before (ref?) and can be used as validation of this task. The re-tracing task has also been used before, but in order to test evoked resets [@Nakayama2019].

## Results

In Figure X, we observe that the data appear to show many resets. We detect them using X, Y, Z algorithm.

[Actual results] For the majority of participants (X/9) we see that most trials show at least one reset (>90%?). In the remainder we still see >50% of the trials have a reset.

The points at which resets appear can be converted to angle relative to the origin of the stimulus, and this should correspond to the [arrow responses]. Do correlation.

It may be that during online tracking, proprioceptive information on the location of the hand is included in the estimate of the location of the stimulus. This may have decreased the likelihood of a reset, and may also changed the strength of the illusion. Since some participants did both experiments (X/9), we can at least test if illusion strength is different for online tracking and offline re-tracing.

# Discussion

We set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli. To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus continuously. While the direction of the tracking movements scaled with the internal motion of the stimulus (was sensitive to the strength of the illusion), there were no clear reset points in the tracking data. In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass. In this re-tracing experiment the drawn traces reflect the strength of the illusion [just] as well, and in most trials there is a clear reset point. The positions of these reset points do no vary much with elapsed time, but are better explained as triggered when the distance between perceived and true position of the stimulus exceeds some threshold. This indicates that while the perceived location of a moving object can be heavily biased by its well-defined, internal motion, an uncertain (probably retinal?) position signal is still used to keep the perceived position of moving objects close to reality.

§3 (one computational model):

- Explain paper with model by Kwon, Tadin & Knill, and VSS poster with fat-tailed noise.

While this model produces reasonable resets, the properties of resets, such as their spatial or temporal distributions have not been characterized yet. An alternative explanation for resets could be that the true but imprecise position that is available for eye-movements can limit or reset the perceived position under some circumstances. In order to get at the properties of resets

If the double-drift illusion is taken as a vector average between the external motion and some proportion of the internal motion, the initial path of the stimulus can be modelled as a straight line from the origin at some angle. This straight line segment should intersect the position of the first reset. We will now see if resets occur at a given distance perpendicular to the path of the real stimulus, after a given amount of time, or perhaps a combination of the two.

This means that it seems that resets occur predominantly at some given discrepancy between the true and illusory position of the stimulus, with a small contribution of time. It may that after seeing the illusion for some time, a ‘source estimation’ mechanism can no longer account for both the perceived, illusory position and the imprecise retinal position, as coming from the same stimulus. Then the visual system changes the position of the stimulus to the position derived from retinotopic information (instead of opting for two separate sources, which is highly unlikely in this case), resulting in a “reset”.


## Conclusion

We find that online tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets. We do find spontaneous resets in single-pass retracing. The illusion appears to resets predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true and illusory position of the stimulus.

# Acknowledgements

The research was support by funding from NSERC Canada to PC and DYPH , as well as by the Department of Psychological and Brain Science, Dartmouth College (PC).

# References {#references .unnumbered}